Contracts for unions and intersections are hard for Nickel

### 2021-06-04 paper.tex
Most of the introduction

\documentclass[sigplan,10pt,review,anonymous]{acmart}
%\usepackage{amssymb}
\usepackage{mathtools,amsmath,amsfonts,amsxtra,amsthm,mathrsfs,mathpartir}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{array}
\usepackage[toc,page]{appendix}
\usepackage{graphicx}
\usepackage{varioref}
\usepackage{verbatim,comment}
\usepackage{longtable}
\usepackage{tikz}
\usepackage{url}
\usepackage{hyperref}

@@

A useful feature of a configuration language is to provide facilities
for schema validation. That is, help answer questions like: does our
configuration have all the required fields? does this field indeed
contains a URL?

These are inherently dynamic questions, as they are all questions
about the evaluated configuration. To this effect, Nickel lets us
annotate any expression with a dynamic schema check: \nickel{exp |
  C}. There is also syntactic sugar to annotate definitions:
\nickel{let id | C = value in exp} stands for \nickel{let id = (value
  | C) in exp}.

@@

But let's consider now writing something like \nickel{nixpkgs |
  packages}, to guarantee that all the packages conform to the desired
schema. If this were a simple Boolean test, it would have to evaluate
all 50\,000 package to check their validity, hence breaking the
laziness of dictionaries. Do we have to choose between laziness and
schema validation? Fortunately, we don't! Enter
\emph{contracts}~\cite{FindlerFelleisenHOContracts}: dynamic check
which can be partially delayed, yet errors can be reported
accurately. Contracts can respect laziness of dictionaries, and they
can be use to add schema validation to functions as well (in fact
functions were their original motivations).

The is no Boolean function which can check that a value has type

@@

This is why the authors have been developing the Nickel
language\cite{NickelRepo}. Nickel is a configuration programming language,
meaning that a Nickel program evaluates to a value that must be
straightforwardly serializable to JSON, YAML or alike. Nickel aims at being
lightweight and easy to write while providing support for writing correct code
and generating valid data. Nickel originated as a rethink of the language of the
Nix package manager, used to describe one of the biggest sets of packaged
software in existence\cite{repology}.  Nickel targets build systems, cloud
deployment tools (Kubernetes, Terraform or NixOps), or anything requiring a
non-trivial configuration.

@@

The end goal being data generation, Nickel is a dynamic, pure and functional
language. Our goal is that writing a simple configuration in Nickel feels as
easy as writing JSON or YAML. Unlike JSON, though, Nickel anticipates large
configurations by being programmable, and by providing means of ensuring
both correctness and validity:

\begin{description}
    \item[(correctness)] The program runs without errors. Errors may be type
        errors, raised for example when evaluating the nonsensical division
        \nickel{1 / "ftp"}. More generally, they can be any the violation of a
        pre- or post-condition. When such an error cannot be prevented, we want
        at least early failures and relevant error messages.\todo{make this
        definition more precise}
    \item[(validity)] The program generates valid data, i.e. data that conforms
        to a schema (imposed by the system consuming the configuration). For
        example, when configuring a web service, the schema may require the
        presence of a field \lstinline+port+ that holds a valid port number.
\end{description}

A traditional answer is to use static types, and in particular static types for
correctness. However, Nickel programs are of a special kind: they are usually
simple, terminating programs that run on fixed inputs. Correctness errors will
show up at evaluation anyway, or will not matter (dead code). With respect to
validation, traditional types are too limited. Validation commonly involves
checking requirements such as being a valid port number or a valid URL which
are, while not technically impossible, really hard to check statically.

In a dynamic language, it is natural and easier to validate data at runtime.
This underlies the approach of dynamic typing. But dynamically typed languages
usually only check the same simple predicates as static types (being a number,
being a string, etc.), and solely for data that are arguments to a primitive
operation. For validation, one needs to check arbitrary predicates on arbitrary
values. This is provided in a basic form by assertions, available in most
languages out there. An assertion is an expression that verifies that a
predicate holds, or abort the program otherwise.  While assertions are fine for
simple sanity checks, their use can incur verbosity, they are clumsy to compose
and to reuse, and they provide limited context attached to an error. It turns
out there is a better tool than plain assertions: contracts.

@@

Assertions are usually mainly used for correctness by enforcing pre- and
post-conditions of functions at runtime. In their foundational
paper\cite{FindlerFelleisenHOContracts}, Findler and Felleisen introduce a
principled approach to run-time assertion checking called contracts. As opposed
to plain assertions, contracts can be composed naturally into more complex ones.
Nickel features such a contract system. For example, contracts nicely support
higher-order functions. If the user has written an \nickel{Even} contract for
numbers, they can use it right away to form a function contract \nickel{Even ->
Int} contract, as illustrated in Figure~\ref{fig:contract-higher-order}. This
contract will check at each call that the argument is an even number
(pre-condition) and that the return value is a integer (post-condition).
Contracts also introduce the notion of blame which is crucial to good error
reporting in presence of higher-order contracts and other composed contracts.

\begin{figure}[h]
  \begin{center}
\begin{lstlisting}[language=Nickel]
let Even = contracts.fromPred (fun val =>
  builtins.isNum val
  && val % 2 == 0) in
let Int = contracts.fromPred (fun val =>
  builtins.isNum val
  && val % 1 == 0) in
{divBy2 | Even -> Int = fun x => x/2}
\end{lstlisting}
  \end{center}
\caption{Higher-order contract}
\label{fig:contract-higher-order}
\end{figure}

In a dynamic language, even more so in a lazy one (see the paragraph on Lazyness
in Section~\ref{sec:feat-lang}), error messages may point to a location far
removed from the offending code. Contracts help by making assumptions on the
values passed at component boundaries explicit (components can be functions,
modules, classes, and so on, depending on language).

Figure~\ref{fig:contract-reporting-wo} and Figure~\ref{fig:contract-reporting-w}
give an example of a situation where adding a simple contract annotation
\nickel{Str -> Str} to a function greatly improves the error message. In the
original program of Figure~\ref{fig:contract-reporting-wo}, the
\nickel{catHosts} function takes a string argument, append it to a list that is
then folded. However, the programmer forgot one parameter when calling to
\hbox{\nickel{makeHost},} thus passing a partially applied function instead of a string
to \nickel{catHosts}. Because this argument is put in a list and folded, the
error is raised later on and points to inside the code of \nickel{catHosts} and
\nickel{makeHost}, which may be totally unknown to the caller.

\begin{figure}[h]
  \begin{center}
\begin{lstlisting}[language=nickel,title={Source program}]
let catHosts = fun last =>
  let hosts = ["foo.com", "bar.org"] in
  lists.fold (fun val acc => val ++ "," ++ acc) hosts last in

\begin{lstlisting}[title={Error message},frame=none,numbers=none]
error: Type error
3 | [(*@{\ldots}@*)] "," ++ acc) hosts last in
  |            ^^^
  | This expression has type Fun,
  | but Str was expected
4 |
5 | let mkHost = fun server ext => [(*@{\ldots}@*)] in
  |              -----------------------
  |              evaluated to this
  = ++, 2nd argument
\end{lstlisting}
  \end{center}
\caption{Error reporting without contract}
\label{fig:contract-reporting-wo}
\end{figure}

By adding a simple annotation to \nickel{catHosts} in
Figure~\ref{fig:contract-reporting-w}, the interpreter fails before even
entering the body of catHosts and is able to deliver a clearer message: the
first parameter of \nickel{catHosts} should be a string but the user provided
the body of the function \nickel{makeHost}. A stack trace is also shown,
including the faulty call to \nickel{catHost}.

\begin{figure}[h]
  \begin{center}
\begin{lstlisting}[language=nickel,title={Added contract}]
let catHosts | Str -> Str = fun last =>
\end{lstlisting}

\begin{lstlisting}[title={Error message with contract},frame=none,numbers=none, basicstyle=\footnotesize\ttfamily]
error: Blame error: contract broken by the caller.
  |  Str -> Str
  |  --- expected type of the argument [(*@{\ldots}@*)]
[(*@{\ldots}@*)]
1 | let catHosts | Str -> Str = fun last =>
  |                ^^^^^^^^^^ bound here
[(*@{\ldots}@*)]
6 | catHosts (makeHost "google")
  | --------------------------- (2) calling <func>
\end{lstlisting}
  \end{center}
\caption{Error reporting with contract}
\label{fig:contract-reporting-w}
\end{figure}

>>>>

\documentclass[sigplan,10pt,review,anonymous]{acmart}
%\usepackage{amssymb}
\usepackage{mathtools,amsmath,amsfonts,amsxtra,amsthm,mathrsfs,mathpartir}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{array}
\usepackage[toc,page]{appendix}
\usepackage{graphicx}
\usepackage{varioref}
\usepackage{verbatim,comment}
\usepackage{longtable}
\usepackage{tikz}
\usepackage{url}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{hyperref}

@@

A useful feature of a configuration language is to provide facilities
for schema validation. That is, help answer questions like: does our
configuration have all the required fields? does the \nickel{url} field indeed
contains a URL?

These are inherently dynamic questions, as they are all questions
about the evaluated configuration. To this effect, Nickel lets us
annotate any expression with a dynamic schema check: \nickel{exp |
  C}. There is also syntactic sugar to annotate definitions:
\nickel{let id | C = value in exp} stands for \nickel{let id = (value
  | C) in exp}.\unsure{This point used to be made, to we want to
  reinsert it? However, Nickel programs are of a special kind: they are usually
simple, terminating programs that run on fixed inputs. Correctness errors will
show up at evaluation anyway, or will not matter (dead code).}

@@

But let's consider now writing something like \nickel{nixpkgs |
  packages}, to guarantee that all the packages conform to the desired
schema. If this were a simple Boolean test, it would have to evaluate
all 50\,000 package to check their validity, hence breaking the
laziness of dictionaries. Do we have to choose between laziness and
schema validation? Fortunately, we don't! Enter
\emph{contracts}~\cite{FindlerFelleisenHOContracts}: dynamic check
which can be partially delayed, yet errors can be reported
accurately. Contracts can respect laziness of dictionaries, and they
can be use to add schema validation to functions as well (in fact
functions were the original motivation for contracts).

There is no Boolean function which can check that a value has type
\nickel{Str -> Str}. Instead, a contract for \nickel{Str -> Str}
checks for each call of the function whether
\begin{enumerate}
\item the argument has type \nickel{Str}, otherwise the caller of the
  function is faulty
\item if so, that the returned value has type \nickel{Str}, otherwise
  the implementation of the function is faulty
\end{enumerate}

Like in the case of lazy dictionaries, the checks are delayed.
Contracts keeps track of whether the caller or the implementation is
at fault for a violation, hence it can report precise error
messages. Compare Figure~\ref{fig:contract-reporting-wo} and
Figure~\ref{fig:contract-reporting-w}: in
Figure~\ref{fig:contract-reporting-wo} an error is reported inside the
\nickel{catHosts} function, but \nickel{catHosts} is, in fact,
correct, as is made clear by Figure~\ref{fig:contract-reporting-wo},
where \nickel{catHosts} is decorated with the \nickel{Str -> Str}
contract, and correctly reports that the caller failed to call
\nickel{catHosts} with a string argument.

\begin{figure*}
  \centering
  \begin{subfigure}[b]{0.48\linewidth}
    \begin{lstlisting}[language=nickel]
let catHosts = fun last =>
  let hosts = ["foo.com", "bar.org"] in
  lists.fold (fun val acc => val ++ "," ++ acc) hosts last in

    \begin{lstlisting}[frame=none,numbers=none, basicstyle=\footnotesize\ttfamily]
error: Type error
3 | [(*@{\ldots}@*)] "," ++ acc) hosts last in
  |            ^^^
  | This expression has type Fun,
  | but Str was expected
4 |
5 | let mkHost = fun server ext => [(*@{\ldots}@*)] in
  |              -----------------------
  |              evaluated to this
  = ++, 2nd argument
\end{lstlisting}
    \caption{Error reporting without contract}
    \label{fig:contract-reporting-wo}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.48\linewidth}
    \begin{lstlisting}[language=nickel]
let catHosts | Str -> Str = fun last =>
  [(*@{\ldots}@*)]
\end{lstlisting}
\vspace{10ex}
    \begin{lstlisting}[frame=none,numbers=none, basicstyle=\footnotesize\ttfamily]
error: Blame error: contract broken by the caller.
  |  Str -> Str
  |  --- expected type of the argument [(*@{\ldots}@*)]
[(*@{\ldots}@*)]
1 | let catHosts | Str -> Str = fun last =>
  |                ^^^^^^^^^^ bound here
[(*@{\ldots}@*)]
6 | catHosts (makeHost "google")
  | --------------------------- (2) calling <func>
    \end{lstlisting}
    \caption{Error reporting with contract}
    \label{fig:contract-reporting-w}
  \end{subfigure}
  \caption{Contracts improve error messages}
\end{figure*}

@@

@@

With respect to
validation, traditional types are too limited. Validation commonly involves
checking requirements such as being a valid port number or a valid URL which
are, while not technically impossible, really hard to check statically.

@@
